Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-26 submission/manuscript.Rmd
shigeki updates fig 5

Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress two hippocampal germline DEGs[@scandagliaLossKdm5cCauses2017]. There is some evidence KDM5C binds to select germline gene promoters in ESCs but is lost in neurons[@scandagliaLossKdm5cCauses2017]. However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters. Thus, it is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], however it is unclear how KDM5C interacts with other suppr.

recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023].

 Somatic repression of germline genes is typically established during the transition between na&#239;ve and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation. In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters, including histone 2A 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Germline genes are then silenced long-term in EpiLCs by _de novo_ placement of DNA CpG methylation (DNAme)[@mochizukiRepressionGermlineGenes2021]. Although KDM5C's catalytic activity is not required to suppress _Dazl_ in ESCs, KDM5C may promote long-term germline gene silencing via H3K4me3 removal since H3K4me3 can impede DNAme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and DNAme is lost at select germline gene promoters in the hippocampus[@scandagliaLossKdm5cCauses2017]. Because KDM5C's role in germline gene repression has only been characterized in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during transition between ESCs and EpiLCs and if its catalytic activity is required for long-term silencing.

@@

We then evaluated KDM5C's embryonic role in germline gene silencing during ESC to EpiLC differentiation. We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation (Figure 5C). While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 5D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 5E). We then perfomed whole genome bislufite sequencing in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs) to determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters. While wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).

- Catalytic activity
- Altogether, this suggests KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes germline gene silencing via removal of H3K4me2/3 during early embryogenesis.

- **note:** should I test if incerase in H3K4me2 and H3K4me3 was highest in _Kdm5c_-KO cells at genes bound by KDM5C at their promoter in EpiLCs? Don't think it's that impactful

>>>>

Previous work suggests KDM5C represses germline genes during early development, as re-expression of KDM5C in knockout neuronal cultures fails to suppress two hippocampal germline DEGs[@scandagliaLossKdm5cCauses2017]. There is some evidence KDM5C binds to select germline gene promoters in embryonic stem cells (ESCs) but is lost in neurons[@scandagliaLossKdm5cCauses2017]. However, the lack of a comprehensive list of germline-enriched genes prohibited systematic characterization of KDM5C binding at germline gene promoters. Thus, it is currently unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

@@

In the early embryo, germline gene promoters are initially decorated with repressive histone modifications before long-term silencing by _de novo_ DNA CpG methylation (CpGme)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Our results above suggest KDM5C also acts during this time period, however how KDM5C interacts with other germline gene silencing mechanism is currently unclear. KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3), marks enriched at active gene promoters[@iwaseXLinkedMentalRetardation2007]. However, KDM5C's catalytic activity was recently shown to not be required for suppressing _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023]. Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required for long-term germline gene silencing. In support of this, CpGme is markedly reduced at two germline gene promoters in the adult hippocampus[@scandagliaLossKdm5cCauses2017]. However, because KDM5C's role in germline gene repression has only been characterized at select germline genes in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during initial CpGme placement.

@@

We then evaluated KDM5C's role in germline gene silencing during ESC to EpiLC differentiation. We first characterized KDM5C's mRNA and protein expression at 0, 24, and 48 hours of differentiation (Figure 5C). While _Kdm5c_ mRNA steadily decreased from 0 to 48 hours of diferentiation (Figure 5D), KDM5C protein initially increased from 0 to 24 hours but then decreased to near knockout levels by 48 hours (Figure 5E). To determine KDM5C's role in the initial placement of DNA methylation at germline gene promoters, we perfomed whole genome bislufite sequencing (WGBS) in wild-type and _Kdm5c_-KO ESCs and 96 hour extended EpiLCs (exEpiLCs). Although wild-type cells accumulated high levels of DNA methylation at germline gene promoters over the course of ESCs to exEpiLC differentiation, DNA methylation was markedly reduced in _Kdm5c_-KO exEpiLCs (Figure 5F).

- Altogether, this suggests KDM5C is dynamically regulated during ESC to EpiLC differentiation and promotes germline gene silencing via removal of H3K4me2/3 during early embryogenesis.
